Component response rate variation underlies the stability of complex systems (revision notes)

### 2020-01-25 notebook/revision_notes.Rmd
More to the argument for a numerical approach, why an analytical one probably won't work

The effect of $\gamma$ does not necessarily result in a circular eigenvalue spectra of $\textbf{M}$. In both panels of Figure 1, $S =$ `r S` and $\sigma_{A} = 1/\sqrt{S}$. In Figures 1a,b and 1c,d, $d = 0$ and $d = 1$, respectively, causing the eigenvalue spectra of $\textbf{A}$ to differ on the real axis (Figure a,c). Figure panels b and c show eigenvalue spectra after $\gamma$ is included such that $M = \gamma A$, where elements of $\gamma$ are drawn from a uniform distribution $\gamma \sim \mathcal{U}(0, 2)$. In both cases, $\sigma^{2}_{A} =$ `r round(sigma, digits = 4)` and $\sigma^{2}_{\gamma} =$ `r round(joint_var, digits = 4)`. Nevertheless, the change in eigenvalue spectra caused by $\gamma$ can affect $\max(\Re(\lambda))$ and therefore system stability.

Stability also decreases with increasing $\rho$ such that $\max(\Re(\lambda)) \approx \sigma_{M}\sqrt{SC}\left(1 + \rho\right)$. Hence, when $\rho$ values are negative, such as when $\textbf{M}$ models a predator-prey system in which $M_{ij}$ and $M_{ji}$ have opposing signs, stability is expected to increase [@Allesina2012]. When elements of $\gamma$ vary independently, the magnitude of $\rho$ is decreased because $\sigma^{2}_{\gamma}$ increases the variance of $M_{ij}$ without affecting the covariance between $M_{ij}$ and $M_{ji}$,

@@

```{r, echo = FALSE, fig.height = 8, fig.width = 6}
sim10 <- read.csv(file = "sim_results/rhos/sim10_rho.csv");
sim15 <- read.csv(file = "sim_results/rhos/sim15_rho.csv");
sim20 <- read.csv(file = "sim_results/rhos/sim20_rho.csv");
sim25 <- read.csv(file = "sim_results/rhos/sim25_rho.csv");
sim30 <- read.csv(file = "sim_results/rhos/sim30_rho.csv");
sim35 <- read.csv(file = "sim_results/rhos/sim35_rho.csv");
par(mar = c(0.25, 0.25, 0.25, 0.25), mfrow = c(3, 2), oma = c(6, 6, 1, 1));
plot(x = sim10[,1], y = sim10[,20], type = "l", pch = 20, lwd = 2, xaxt = "n",
     ylim = c(-1, 3.5), cex.axis = 1.5);
points(x = sim10[,15], y = sim10[,21], type = "l", lwd = 2, lty = "dashed");
abline(h = 0, col = "red", lty = "dotted", lwd = 0.8);
text(x = -0.6, y = 3.2, labels = "S = 10", cex = 2.5);
plot(x = sim15[,1], y = sim15[,20], type = "l", pch = 20, lwd = 2, xaxt = "n",
     ylim = c(-1, 3.5), yaxt = "n", cex.axis = 1.5);
points(x = sim15[,15], y = sim15[,21], type = "l", lwd = 2, lty = "dashed");
abline(h = 0, col = "red", lty = "dotted", lwd = 0.8);
text(x = -0.6, y = 3.2, labels = "S = 15", cex = 2.5);
plot(x = sim20[,1], y = sim20[,20], type = "l", pch = 20, lwd = 2, xaxt = "n",
     ylim = c(-1, 3.5), cex.axis = 1.5);
points(x = sim20[,15], y = sim20[,21], type = "l", lwd = 2, lty = "dashed");
abline(h = 0, col = "red", lty = "dotted", lwd = 0.8);
text(x = -0.6, y = 3.2, labels = "S = 20", cex = 2.5);
plot(x = sim25[,1], y = sim25[,20], type = "l", pch = 20, lwd = 2, xaxt = "n",
     ylim = c(-1, 3.5), yaxt = "n", cex.axis = 1.25);
points(x = sim25[,15], y = sim25[,21], type = "l", lwd = 2, lty = "dashed");
abline(h = 0, col = "red", lty = "dotted", lwd = 0.8);
text(x = -0.6, y = 3.2, labels = "S = 25", cex = 2.5);
plot(x = sim30[,1], y = sim30[,20], type = "l", pch = 20, lwd = 2, 
     ylim = c(-1, 3.5), cex.axis = 1.5);
points(x = sim30[,15], y = sim30[,21], type = "l", lwd = 2, lty = "dashed");
abline(h = 0, col = "red", lty = "dotted", lwd = 0.8);
text(x = -0.6, y = 3.2, labels = "S = 30", cex = 2.5);
plot(x = sim35[,1], y = sim35[,20], type = "l", pch = 20, lwd = 2, 
     ylim = c(-1, 3.5), yaxt = "n", cex.axis = 1.5);
points(x = sim35[,15], y = sim35[,21], type = "l", lwd = 2, lty = "dashed");
abline(h = 0, col = "red", lty = "dotted", lwd = 0.8);
text(x = -0.6, y = 3.2, labels = "S = 35", cex = 2.5);
mtext(text =  expression(paste("E Correlation between A"[ij]," and A"[ji], 
                             " (",rho,")")), side = 1,
      line = 4, outer = TRUE, cex = 1.5);
mtext(text = "E leading real part eigenvalue", side = 2,
      line = 3.5, outer = TRUE, cex = 1.5);
```

>>>>

The effect of $\gamma$ does not necessarily result in a circular eigenvalue spectra of $\textbf{M}$. In both panels of Figure 1, $S =$ `r S` and $\sigma_{A} = 1/\sqrt{S}$. In Figures 1a, b and 1c, d, $d = 0$ and $d = 1$, respectively, causing the eigenvalue spectra of $\textbf{A}$ to differ on the real axis (Figure a,c). Figure panels b and c show eigenvalue spectra after $\gamma$ is included such that $\textbf{M} = \gamma \textbf{A}$, where elements of $\gamma$ are drawn from a uniform distribution $\gamma \sim \mathcal{U}(0, 2)$. In both cases, $\sigma^{2}_{A} =$ `r round(sigma, digits = 4)` and $\sigma^{2}_{\gamma} =$ `r round(joint_var, digits = 4)`. Nevertheless, the change in eigenvalue spectra caused by $\gamma$ can affect $\max(\Re(\lambda))$ and therefore system stability.

Stability also decreases with increasing $\rho$ such that $\max(\Re(\lambda)) \approx \sigma_{M}\sqrt{SC}\left(1 + \rho\right)$. When $\rho$ values are negative, such as when $\textbf{M}$ models a predator-prey system in which $M_{ij}$ and $M_{ji}$ have opposing signs, stability is expected to increase [@Allesina2012]. When elements of $\gamma$ vary independently, the magnitude of $\rho$ is decreased because $\sigma^{2}_{\gamma}$ increases the variance of $M_{ij}$ without affecting the covariance between $M_{ij}$ and $M_{ji}$,

@@

```{r, echo = FALSE, fig.height = 8, fig.width = 6}
sim10 <- read.csv(file = "sim_results/rhos/sigma0pt2/sim10.csv");
sim15 <- read.csv(file = "sim_results/rhos/sigma0pt2/sim15.csv");
sim20 <- read.csv(file = "sim_results/rhos/sigma0pt2/sim20.csv");
sim25 <- read.csv(file = "sim_results/rhos/sigma0pt2/sim25.csv");
sim30 <- read.csv(file = "sim_results/rhos/sigma0pt2/sim30.csv");
sim35 <- read.csv(file = "sim_results/rhos/sigma0pt2/sim35.csv");
sim40 <- read.csv(file = "sim_results/rhos/sigma0pt2/sim40.csv");
sim45 <- read.csv(file = "sim_results/rhos/sigma0pt2/sim45.csv");
par(mar = c(0.25, 0.25, 0.25, 0.25), mfrow = c(3, 2), oma = c(6, 6, 1, 1));
plot(x = sim20[,1], y = sim20[,20], type = "l", pch = 20, lwd = 2, xaxt = "n",
     ylim = c(-1, 3.5), cex.axis = 1.5);
points(x = sim20[,15], y = sim20[,21], type = "l", lwd = 2, lty = "dashed");
abline(h = 0, col = "red", lty = "dotted", lwd = 0.8);
text(x = -0.6, y = 3.2, labels = "S = 20", cex = 2.5);
plot(x = sim25[,1], y = sim25[,20], type = "l", pch = 20, lwd = 2, xaxt = "n",
     ylim = c(-1, 3.5), yaxt = "n", cex.axis = 1.5);
points(x = sim25[,15], y = sim25[,21], type = "l", lwd = 2, lty = "dashed");
abline(h = 0, col = "red", lty = "dotted", lwd = 0.8);
text(x = -0.6, y = 3.2, labels = "S = 25", cex = 2.5);
plot(x = sim30[,1], y = sim30[,20], type = "l", pch = 20, lwd = 2, xaxt = "n",
     ylim = c(-1, 3.5), cex.axis = 1.5);
points(x = sim30[,15], y = sim30[,21], type = "l", lwd = 2, lty = "dashed");
abline(h = 0, col = "red", lty = "dotted", lwd = 0.8);
text(x = -0.6, y = 3.2, labels = "S = 30", cex = 2.5);
plot(x = sim35[,1], y = sim35[,20], type = "l", pch = 20, lwd = 2, xaxt = "n",
     ylim = c(-1, 3.5), yaxt = "n", cex.axis = 1.25);
points(x = sim35[,15], y = sim35[,21], type = "l", lwd = 2, lty = "dashed");
abline(h = 0, col = "red", lty = "dotted", lwd = 0.8);
text(x = -0.6, y = 3.2, labels = "S = 35", cex = 2.5);
plot(x = sim40[,1], y = sim40[,20], type = "l", pch = 20, lwd = 2, 
     ylim = c(-1, 3.5), cex.axis = 1.5);
points(x = sim40[,15], y = sim40[,21], type = "l", lwd = 2, lty = "dashed");
abline(h = 0, col = "red", lty = "dotted", lwd = 0.8);
text(x = -0.6, y = 3.2, labels = "S = 40", cex = 2.5);
plot(x = sim45[,1], y = sim45[,20], type = "l", pch = 20, lwd = 2, 
     ylim = c(-1, 3.5), yaxt = "n", cex.axis = 1.5);
points(x = sim45[,15], y = sim45[,21], type = "l", lwd = 2, lty = "dashed");
abline(h = 0, col = "red", lty = "dotted", lwd = 0.8);
text(x = -0.6, y = 3.2, labels = "S = 45", cex = 2.5);
mtext(text =  expression(paste("E Correlation between A"[ij]," and A"[ji], 
                               " (",rho,")")), side = 1,
      line = 4, outer = TRUE, cex = 1.5);
mtext(text = "E leading real part eigenvalue", side = 2,
      line = 3.5, outer = TRUE, cex = 1.5);
```

### 2020-01-25 notebook/revision_notes.tex
More to the argument for a numerical approach, why an analytical one probably won't work

\section{Effects of correlations
interactions}\label{effects-of-correlations-interactions}

@@

The effect of \(\gamma\) does not necessarily result in a circular
eigenvalue spectra of \(\textbf{M}\). In both panels of Figure 1,
\(S =\) 400 and \(\sigma_{A} = 1/\sqrt{S}\). In Figures 1a and 1b,
\(d = 0\) and \(d = 1\), respectively, causing the eigen spectra of
\(\textbf{A}\) to differ on the real axis (blue points). Red points she
eigen spectra after \(\gamma\) is include such that \(M = \gamma A\),
where elements of \(\gamma\) are drawn from a uniform distribution
\(\gamma \sim \mathcal{U}(0, 2)\). In both cases, \(\sigma^{2}_{A} =\)
0.05 and \(\sigma^{2}_{\gamma} =\) 0.0033.

\section{More down}\label{more-down}

@@

Note that the correlation \(\rho\) adjusts the criteria for stability as
follows\textsuperscript{\protect\hyperlink{ref-Allesina2015a}{2}},

@@

Hence, under the same set of parameters, we predict a linear effect of
\(\rho\) on \(\max(\Re(\lambda)) = 0\), such that
\(\max(\Re(\lambda)) = \rho\). Allesina and Tang note that \(\rho\) is
defined as
follows\textsuperscript{\protect\hyperlink{ref-Allesina2015a}{2}},

@@

Gibbs et al.\textsuperscript{\protect\hyperlink{ref-Gibbs2017}{3}}
showed analytically that as \(S \to \infty\), the effect of a diagonal
matrix \(\textbf{X}\) on the stability of \(\textbf{M} = \textbf{XA}\)
becomes negligible. They also note that local stability of a random
matrix can be predicted from simple moments affecting three parameters,

@@

When \(\textbf{A}\) was stabilised by \(Var(\gamma)\), the change in
\(\rho\) was normally distributed around 4.8394472\(\times 10^{-6}\)
(95\% CIs: -74.3396367\(\times 10^{-6}\), 77.6796914\(\times 10^{-6}\)).
Hence, decreasing the correlation between \(\textbf{A}_{ij}\) and
\(\textbf{A}_{ji}\) was not by itself the cause of stability. For a
clearer picture of the effect of \(\gamma\), it is useful to show the
relationship between \(\rho\) and \(\max(\Re(\lambda))\) again as above,
but this time also for how the relationship changes given
\(Var(\gamma)\). That is, given \(\textbf{M} = \textbf{A}\) in addition
to \(\textbf{M} = \gamma\textbf{A}\).

@@

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\hypertarget{ref-Tang2014b}{}
1. Tang, S. \& Allesina, S. Reactivity and stability of large
ecosystems. \emph{Frontiers in Ecology and Evolution} \textbf{2,} 1--8
(2014).

\hypertarget{ref-Allesina2015a}{}
2. Allesina, S. \& Tang, S. The stability--complexity relationship at
age 40: a random matrix perspective. \emph{Population Ecology} 63--75
(2015).
doi:\href{https://doi.org/10.1007/s10144-014-0471-0}{10.1007/s10144-014-0471-0}

\hypertarget{ref-Gibbs2017}{}
3. Gibbs, T., Grilli, J., Rogers, T. \& Allesina, S. The effect of
population abundances on the stability of large random ecosystems.
\emph{Physical Review E - Statistical, Nonlinear, and Soft Matter
Physics} \textbf{98,} 022410 (2018).

>>>>

\hypertarget{effects-of-correlations-interactions}{%
\section{Effects of correlations
interactions}\label{effects-of-correlations-interactions}}

@@

The effect of \(\gamma\) does not necessarily result in a circular
eigenvalue spectra of \(\textbf{M}\). In both panels of Figure 1,
\(S =\) 400 and \(\sigma_{A} = 1/\sqrt{S}\). In Figures 1a,b and 1c,d,
\(d = 0\) and \(d = 1\), respectively, causing the eigenvalue spectra of
\(\textbf{A}\) to differ on the real axis (Figure a,c). Figure panels b
and c show eigenvalue spectra after \(\gamma\) is included such that
\(M = \gamma A\), where elements of \(\gamma\) are drawn from a uniform
distribution \(\gamma \sim \mathcal{U}(0, 2)\). In both cases,
\(\sigma^{2}_{A} =\) 0.05 and \(\sigma^{2}_{\gamma} =\) 0.0031.
Nevertheless, the change in eigenvalue spectra caused by \(\gamma\) can
affect \(\max(\Re(\lambda))\) and therefore system stability.

Stability also decreases with increasing \(\rho\) such that
\(\max(\Re(\lambda)) \approx \sigma_{M}\sqrt{SC}\left(1 + \rho\right)\).
Hence, when \(\rho\) values are negative, such as when \(\textbf{M}\)
models a predator-prey system in which \(M_{ij}\) and \(M_{ji}\) have
opposing signs, stability is expected to
increase\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}}. When
elements of \(\gamma\) vary independently, the magnitude of \(\rho\) is
decreased because \(\sigma^{2}_{\gamma}\) increases the variance of
\(M_{ij}\) without affecting the covariance between \(M_{ij}\) and
\(M_{ji}\),

\[\rho = \frac{cov(M_{ij}, M_{ji})}{\sigma^{2}_{A}(1 + \sigma^{2}_{\gamma})}.\]
Because \(\gamma\) does not affect \(cov(M_{ij}, M_{ji})\), the sign of
\(\rho\) remains unchanged.

\hypertarget{more-down}{%
\section{More down}\label{more-down}}

@@

Note that the correlation \(\rho\) adjusts the criteria for stability as
follows\textsuperscript{\protect\hyperlink{ref-Allesina2015a}{3}},

@@

Hence, under the same set of parameters, we predict a linear effect of
\(\rho\) on \(\max(\Re(\lambda)) = 0\), such that
\(\max(\Re(\lambda)) = \rho\). Allesina and Tang note that \(\rho\) is
defined as
follows\textsuperscript{\protect\hyperlink{ref-Allesina2015a}{3}},

@@

Gibbs et al.\textsuperscript{\protect\hyperlink{ref-Gibbs2017}{4}}
showed analytically that as \(S \to \infty\), the effect of a diagonal
matrix \(\textbf{X}\) on the stability of \(\textbf{M} = \textbf{XA}\)
becomes negligible. They also note that local stability of a random
matrix can be predicted from simple moments affecting three parameters,

@@

When \(\textbf{A}\) was stabilised by \(Var(\gamma)\), the change in
\(\rho\) was normally distributed around 4.8394472\(\times 10^{-6}\)
(95\% CIs: -75.9050127\(\times 10^{-6}\), 84.5804827\(\times 10^{-6}\)).
Hence, decreasing the correlation between \(\textbf{A}_{ij}\) and
\(\textbf{A}_{ji}\) was not by itself the cause of stability. For a
clearer picture of the effect of \(\gamma\), it is useful to show the
relationship between \(\rho\) and \(\max(\Re(\lambda))\) again as above,
but this time also for how the relationship changes given
\(Var(\gamma)\). That is, given \(\textbf{M} = \textbf{A}\) in addition
to \(\textbf{M} = \gamma\textbf{A}\).

@@

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-Tang2014b}{}%
1. Tang, S. \& Allesina, S. Reactivity and stability of large
ecosystems. \emph{Frontiers in Ecology and Evolution} \textbf{2,} 1--8
(2014).

\leavevmode\hypertarget{ref-Allesina2012}{}%
2. Allesina, S. \& Tang, S. Stability criteria for complex ecosystems.
\emph{Nature} \textbf{483,} 205--208 (2012).

\leavevmode\hypertarget{ref-Allesina2015a}{}%
3. Allesina, S. \& Tang, S. The stability--complexity relationship at
age 40: a random matrix perspective. \emph{Population Ecology} 63--75
(2015).
doi:\href{https://doi.org/10.1007/s10144-014-0471-0}{10.1007/s10144-014-0471-0}

\leavevmode\hypertarget{ref-Gibbs2017}{}%
4. Gibbs, T., Grilli, J., Rogers, T. \& Allesina, S. The effect of
population abundances on the stability of large random ecosystems.
\emph{Physical Review E - Statistical, Nonlinear, and Soft Matter
Physics} \textbf{98,} 022410 (2018).
